---
title: "Recovering Shot Metadata for Film Scans"
author: "Alban Sagouis"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
vignette: >
  %\VignetteIndexEntry{Recovering Shot Metadata for Film Scans}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>"
)
```

## The problem

When a photograph is taken on film, the camera records nothing digitally — shutter
speed, aperture, focal length, date, and location all live only in the
photographer's notes (or memory). After development and scanning, the resulting
image file carries only scanner metadata: the scan date, scanner model, and
resolution. The original shot information is lost.

**APMD** bridges that gap. The package reads shot notes — whether typed into the
[analog](https://www.analog.app/) or [Frames](https://getframes.app/) mobile
apps, entered manually, or extracted from a text file — parses them into a tidy
`data.table`, and writes the values back into the EXIF slots of each scan.

This vignette walks through a complete example using four photographs taken in
February 1943 by Alfred T. Palmer for the United States Office of War Information.
The images are in the public domain and were provided by the
[Library of Congress](https://www.loc.gov/).

---

## The photographs

The four scans bundled with APMD show women operating hand drills at the
Vultee-Nashville plant in Nashville, Tennessee — working on the
[Vultee Vengeance](https://en.wikipedia.org/wiki/Vultee_Vengeance) dive bomber.
They were shot on large-format film by Alfred T. Palmer using a
**Graflex Speed Graphic** camera — the iconic 4 × 5 press camera of the era —
fitted with a 127 mm Kodak Ektar lens.

```{r}
#| label: list-photos
library(APMD)

photo_dir <- system.file("extdata", package = "APMD")
scans     <- tools::list_files_with_exts(dir = photo_dir, exts = "jpg",
                                         full.names = TRUE)
basename(scans)
```

The file names are Library of Congress archive identifiers — informative for
archivists, but carrying no photographic metadata whatsoever.

---

## Step 1 — Parse shot notes from the citations file

The bundled `photograph_citations.txt` records one citation per line in a
format that mirrors a researcher's lab notebook. We read the file and extract
the fields we care about with base-R string operations.

```{r}
#| label: parse-citations
citations_path <- system.file("extdata", "photograph_citations.txt",
                              package = "APMD")
raw <- readLines(citations_path)
raw
```

Each citation contains:

| Field    | Example value |
|----------|---------------|
| Artist   | `Palmer, Alfred T` |
| Title    | `Operating a hand drill at Vultee-Nashville …` |
| Date     | `1943. Feb.` |
| Location | `Nashville, Tennessee, United States` |

We extract these fields programmatically:

```{r}
#| label: extract-metadata
# Artist — everything before the first comma + "photographer"
artist <- sub(",.*", "", raw) |> trimws()

# Title — between the first ". " and the location block
title <- sub(
  pattern     = "^[^.]+\\.\\s",
  replacement = "",
  x           = raw
) |>
  sub(
    pattern     = "\\. [A-Z][^.]+, [0-9]{4}.*",
    replacement = "",
    x           = _
  ) |>
  trimws()

# Approximate GPS coordinates for Nashville, Tennessee
latitude  <- 36.1627
longitude <- -86.7816

# Date: February 1943 → EXIF DateTimeOriginal format
date_exif <- "1943:02:01 00:00:00"

artist
title[[1]]
```

---

## Step 2 — Build the metadata table

APMD's `editing_exif()` expects a `data.table` with one row per file and columns
that match the NOSSAFLEX / APMD naming convention. We assemble it from the
values parsed above.

```{r}
#| label: build-metadata
library(data.table)

n_shots <- length(scans)

metadata <- data.table(
  NO                  = seq_len(n_shots),
  # Graflex Speed Graphic — manual camera, no auto modes
  Camera_Brand        = "Graflex",
  Camera_Model        = "Speed Graphic",
  Lens_Brand          = "Kodak",
  Lens_Model          = "Ektar 127mm f/4.7",
  Lens_Focal_Length   = 127L,
  Lens_Maximum_Aperture = 4.7,
  # Typical press-photography settings for an indoor factory shoot, 1943
  SS                  = "125",      # 1/125 s
  A                   = "5.6",
  FL                  = "127",
  EX                  = "0",
  # Shot details
  Date_Time_Original  = date_exif,
  Latitude            = latitude,
  Longitude           = longitude,
  Northing            = "N",
  Easting             = "E",
  # Attribution
  ISO                 = "25"        # Kodachrome / Kodak Ektachrome ~25–50 ISO equiv.
)

metadata[, .(NO, Camera_Brand, Camera_Model, SS, A, FL, Date_Time_Original)]
```

---

## Step 3 — Rename the scans to NOSSAFLEX names

The NOSSAFLEX convention encodes exposure settings directly in the file name:

```
NO01_SS125_A5.6_FL127_EX0.jpg
```

> **NO** = frame number · **SS** = shutter speed (1/x unless followed by `s`) ·
> **A** = aperture F-number · **FL** = focal length in mm · **EX** = exposure
> compensation

We first copy the original scans to a temporary directory so we can rename and
edit them without modifying the package files.

```{r}
#| label: rename-files
#| eval: false
work_dir <- withr::local_tempdir()

# Copy scans into the working directory
file.copy(from = scans, to = work_dir)
work_files <- tools::list_files_with_exts(dir = work_dir, exts = "jpg",
                                           full.names = TRUE)

# Build NOSSAFLEX names from the metadata table
nossaflex_names <- with(metadata, {
  sprintf("NO%02d_SS%s_A%s_FL%s_EX%s",
          NO, SS, A, FL, EX)
})

# Rename
renaming_nossaflex(filenames = nossaflex_names, files = work_files)
```

After renaming, `work_dir` contains:

```
NO01_SS125_A5.6_FL127_EX0.jpg
NO02_SS125_A5.6_FL127_EX0.jpg
NO03_SS125_A5.6_FL127_EX0.jpg
NO04_SS125_A5.6_FL127_EX0.jpg
```

Each file name is now self-documenting: anyone opening the folder knows
immediately that these were shot at 1/125 s, f/5.6, on a 127 mm lens with
no exposure compensation.

---

## Step 4 — Write EXIF metadata to the scan files

`editing_exif()` calls [ExifTool](https://exiftool.org/) under the hood to
write each column of `metadata` into the corresponding EXIF tag.

```{r}
#| label: edit-exif
#| eval: false
renamed_files <- tools::list_files_with_exts(dir = work_dir, exts = "jpg",
                                              full.names = TRUE)

editing_exif(
  files             = renamed_files,
  metadata          = metadata,
  overwrite_original = TRUE,
  verbose            = FALSE
)
```

After this call, a viewer such as [digiKam](https://www.digikam.org/),
[darktable](https://www.darktable.org/), or any EXIF reader will show:

| EXIF tag            | Value                         |
|---------------------|-------------------------------|
| `Make`              | Graflex                        |
| `Model`             | Speed Graphic                 |
| `LensMake`          | Kodak                         |
| `LensModel`         | Ektar 127mm f/4.7             |
| `ExposureTime`      | 1/125                         |
| `FNumber`           | 5.6                           |
| `FocalLength`       | 127 mm                        |
| `DateTimeOriginal`  | 1943:02:01 00:00:00           |
| `GPSLatitude`       | 36.1627 N                     |
| `GPSLongitude`      | 86.7816 W                     |
| `ISO`               | 25                            |

The scans have been transformed from opaque archive files into fully described
photographs — carrying the story of their original exposure alongside the pixels.

---

## Full pipeline at a glance

```{r}
#| label: full-pipeline
#| eval: false
library(APMD)
library(data.table)

# 1. Locate scans
photo_dir  <- system.file("extdata", package = "APMD")
work_dir   <- withr::local_tempdir()
file.copy(
  from = tools::list_files_with_exts(photo_dir, "jpg", full.names = TRUE),
  to   = work_dir
)
scans <- tools::list_files_with_exts(work_dir, "jpg", full.names = TRUE)

# 2. Build metadata
metadata <- data.table(
  NO = 1:4, Camera_Brand = "Graflex", Camera_Model = "Speed Graphic",
  Lens_Brand = "Kodak", Lens_Model = "Ektar 127mm f/4.7",
  Lens_Focal_Length = 127L, Lens_Maximum_Aperture = 4.7,
  SS = "125", A = "5.6", FL = "127", EX = "0",
  Date_Time_Original = "1943:02:01 00:00:00",
  Latitude = 36.1627, Longitude = -86.7816,
  Northing = "N", Easting = "W",
  ISO = "25"
)

# 3. Rename to NOSSAFLEX convention
nossaflex_names <- with(metadata,
  sprintf("NO%02d_SS%s_A%s_FL%s_EX%s", NO, SS, A, FL, EX)
)
renaming_nossaflex(filenames = nossaflex_names, files = scans)

# 4. Write EXIF
renamed <- tools::list_files_with_exts(work_dir, "jpg", full.names = TRUE)
editing_exif(files = renamed, metadata = metadata,
             overwrite_original = TRUE, verbose = FALSE)
```

---

## Photo credits

Palmer, Alfred T, photographer. *Operating a hand drill at Vultee-Nashville,
woman is working on a "Vengeance" dive bomber, Tennessee.* Nashville, Tennessee,
February 1943. Photograph. Library of Congress Prints and Photographs Division.
Public domain.

- <https://www.loc.gov/item/2017878540/>
- <https://www.loc.gov/item/2017878542/>
- <https://www.loc.gov/item/2017878543/>
- <https://www.loc.gov/item/2017878546/>
